Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-26 paper/paper.tex
put back the best size of the figures
removed commented out and rewritten fragments to clean up latex.
new style

\documentclass[preprint]{sigplanconf}

\usepackage{siunitx}
\usepackage[color]{changebar}
\usepackage{enumitem}
\usepackage[lighttt]{lmodern}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphicx}
%\usepackage{microtype}

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican,basicstyle=\linespread{1.0}\ttfamily, columns=flexible, showstringspaces=false}

\usepackage{hyperref}
\usepackage{url}
\usepackage{etoolbox}

@@

\title{Design and Implementation of Probabilistic Programming Language Anglican}
\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}

@@

\begin{document}

@@

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
    \item Design and implementation of a probabilistic programming language
        Anglican involving tight bilateral integration with a general-purpose
        programming language.
    \item Techniques for efficient and compact implementation of inference
        algorithms, such as representation of inference results as a lazy
        sequence of samples and a novel scheme for addressing of checkpoints.
    \item Anglican's novel representation of random processes 
            from statistics and machine learning, such as beta-Bernoulli
            process. The representation is
            \emph{stateless} and seamlessly integrates into pure functional 
            computation. This contrasts with typical \emph{stateful}
            implementations of these processes in other probabilistic 
            programming languages.
%    \item Specification of probabilistic forms in Anglican which facilitates 
%        separation of probabilistic programs from inference algorithms.
\end{itemize}

@@

% Inference algorithms exploit the CPS structure of the code to
% evaluate \texttt{sample} and \texttt{observe} checkpoints in an algorithm-specific manner

% \iftoggle{full}{%
% In Anglican and similar languages there are two types
% of such checkpoints:

% \begin{itemize}
%     \item drawing a value from a random source (\texttt{sample});
%     \item conditioning a computed value on a random source
%         (\texttt{observe}).
% \end{itemize}

% }{%
% }%
\iftoggle{full}{%
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle (corresponding to
a user-level thread executing a program)
can be either replicated or terminated. Similarly, in 
Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.

}{%
}%
% Because of the checkpoints, Anglican programs must allow the
% inference algorithm to step in, recording information and
% affecting control flow. This can be implemented through
% coroutines/cooperative multitasking, and parallel execution/preemptive
% multitasking, as well as through explicit
% maintenance of program continuations at checkpoints. Anglican
% follows the latter option.
\iftoggle{full}{%
Clojure is a functional language, and continuation-passing style
(CPS) transformation is a well-developed technique in the area
of functional languages. Implementing a variant of CPS
transformation seemed to be the most flexible and lightweight
option --- any other form of concurrency would put a higher
burden on the underlying runtime (JVM) and the operating system.
Consequently, Anglican has been implemented as a CPS-transformed
computation with access to continuations in probabilistic
checkpoints. 
}{%
}%

@@

%Operational semantics of Anglican queries is different from that
%of Clojure code, therefore queries must be called through
%inference algorithms, rather than `directly'. 
The operational semantics of an Anglican query are different from those
of Clojure code, and are determined by an inference algorithm. Thus,
Anglican queries must be called through these inference algorithms, rather
rather than `directly'. For this purpose,
\iftoggle{full}{the
\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
function \texttt{infer} using Clojure's multimethod mechanism.
This function accepts and runs an Anglican query, 
and returns a lazy sequence of weighted samples from the
distribution defined by the query. Providing an implementation of this function
is a responsibility of the inference algorithm, which should also
override the polymorphic function \texttt{checkpoint} (defined
as a multimethod) so as to handle \texttt{sample} and
\texttt{observe} in an algorithm-specific manner and
to construct an appropriate result on the
termination of a probabilistic program.

%When inference
%is performed on an Anglican query, the query is run by a
%particular inference algorithm. Inference algorithms must
%provide an implementation for \texttt{infer}, as well as
%override the polymorphic function \texttt{checkpoint} (defined
%as a multimethod) so as to handle \texttt{sample} and
%\texttt{observe} in an algorithm-specific manner and
%to construct an appropriate result on the
%termination of a probabilistic program.

@@

%\section{Probabilistic Program}
%\label{sec:pp-def}
%
%Anglican is a language for expressing probabilistic programs.
%A probabilistic program is a stateful \textit{deterministic} computation
%$\mathcal{P}$ with the following properties~\cite{TMP+15}:
%\begin{itemize}
%\item Initially, $\mathcal{P}$ expects no arguments.
%\item On every call, $\mathcal{P}$ returns either a distribution
%  and an address $(F,\alpha)$, a distribution and a value $(G,
%        y)$, or a value $z$.
%\item Upon returning $F$, $\mathcal{P}$ expects a value $x$ drawn from $F$ as
%    the argument to the next call (\texttt{sample} in Anglican).
%\item Upon returning $(G, y)$, $\mathcal{P}$ is invoked again
%    without arguments (\texttt{observe} in Anglican).
%\item Upon returning $z$, $\mathcal{P}$ terminates.
%\end{itemize}
%A program is run by calling $\mathcal{P}$ repeatedly until
%termination.
%
%A program need not generate the same sequence of random variables in every
%execution. For this reason we assume that each random variable $x$ is assigned
%a unique label $\alpha$, which we call an address, that induces a
%correspondence between variables in different executions. Every execution
%implicitly produces a sequence of triples of distributions, values of
%\textit{latent} random variables, and addresses, $(F_i, x_i, \alpha_i)$.  We
%call this sequence a \textit{trace} and denote it by $\pmb{x}$. A trace induces
%a sequence of pairs $(G_j, y_j)$ of distributions and values of
%\textit{observed} random variables. We call this sequence an \textit{image} and
%denote it by $\pmb{y}$. For notational simplicity we assume that the program
%always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.
%
%The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
%is defined in terms of the product of the probabilities of all random
%choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
%\begin{align}
%  \gamma(\pmb{x}) 
%  &:= 
%%  \prod_{i=1}^{\left|\pmb{x}\right|}
%  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
%  \label{eqn:p-trace}
%\end{align}
%The objective of inference in probabilistic program $\mathcal{P}$
%is to discover the distribution of $\pmb{z}$.

@@

%Their operational semantics can be defined in terms of
%importance sampling(Equation~\ref{eqn:p-trace}): 
%\begin{itemize}
%    \item \texttt{sample} returns a value drawn from
%        the distribution.
%    \item \texttt{observe} multiplies the unnormalized
%        probability of the program trace by the probability of
%        drawing the value from the distribution.
%\end{itemize}
%However, the actual computation involved in handling of
%\texttt{sample} and \texttt{observe} depends on the inference
%algorithm. Different inference algorithms may treat
%\texttt{sample} and \texttt{observe} differently as long as they
%compute the same distribution of program results.
%Denotational and operational semantics for an idealized version
%of Anglican is introduced in \cite{SYH+16}; in general though,
%specifying semantics of higher-order probabilistic programming
%languages is an open problem.

@@

;; invoked when algo parameter is :importance
(defmethod infer :importance 
  [algo prog value & {}]
  (letfn
    ;; recursive function without any parameter
    [sample-seq 
      []
      ;; lazy infinite sequence 
      (lazy-seq
        (cons
          (:state (exec ::algorithm prog value
                        initial-state))
          (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
This implementation is dispatched 
when \texttt{infer} receives \texttt{:importance} as its \texttt{algo} 
parameter. (Anglican includes multiple implementations of \texttt{infer}
for different inference algorithms.) Once dispatched, it lazily 
constructs an infinite sequence of inference states by repeatedly 
executing the program \texttt{prog} using \texttt{exec}
and retrieving the final inference state of the execution 
using \texttt{:state}. For checkpoint handlers, importance sampling simply
relies on their default implementations.
%\texttt{anglican.importance/{\linebreak[0]}infer} just calls
%\texttt{anglican.inference/{\linebreak[0]}exec} and relies on default
%implementations of checkpoint handlers.

@@

\acks

@@

\bibliographystyle{abbrvnat}
\bibliography{refs}

>>>>

\documentclass[sigconf]{acmart}

\setcopyright{rightsretained}

\usepackage{siunitx}
\usepackage[color]{changebar}
\usepackage{enumitem}
\usepackage[lighttt]{lmodern}
\usepackage{graphicx}

\usepackage{listings}

\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican,basicstyle=\ttfamily, columns=flexible, showstringspaces=false}

\usepackage{url}
\usepackage{etoolbox}

@@

@@

\acmConference[IFL'16]{Implementation and Application of Functional Programming Languages}{August 2016}{Leuven, Belgium}
\acmYear{2016}
\copyrightyear{2017}

\begin{document}
\title{Design and Implementation of Probabilistic Programming Language Anglican}

\author{David Tolpin}
\affiliation{\institution{University of Oxford}}
\email{dtolpin@robots.ox.ac.uk}

\author{Jan-Willem van de Meent}
\affiliation{\institution{University of Oxford}}
\email{jwvdm@robots.ox.ac.uk}

\author{Hongseok Yang}
\affiliation{\institution{University of Oxford}}
\email{hongseok.yang@cs.ox.ac.uk}

\author{Frank Wood}
\affiliation{\institution{University of Oxford}}
\email{fwood@cs.ox.ac.uk}

@@

\subsection*{Contributions} This paper brings the following
major contributions:
\begin{itemize}
    \item Design and implementation of a probabilistic programming language
        Anglican involving tight bilateral integration with a general-purpose
        programming language.
    \item Techniques for efficient and compact implementation of inference
        algorithms, such as representation of inference results as a lazy
        sequence of samples and a novel scheme for addressing of checkpoints.
    \item Anglican's novel representation of random processes 
            from statistics and machine learning, such as beta-Bernoulli
            process. The representation is
            \emph{stateless} and seamlessly integrates into pure functional 
            computation. This contrasts with typical \emph{stateful}
            implementations of these processes in other probabilistic 
            programming languages.
\end{itemize}

@@

\iftoggle{full}{%
Depending on the \textit{inference algorithm}, \texttt{sample}
and \texttt{observe} may result in implicit input/output
operations and control changes. For example, \texttt{observe} in
particle filtering inference algorithms \cite{WVM14} is a non-deterministic
control statement at which a particle (corresponding to
a user-level thread executing a program)
can be either replicated or terminated. Similarly, in 
Metropolis-Hastings \cite{WSG11},
\texttt{sample} is both an input
statement which `reads' values from a random source,
and a non-deterministic control statement
(with delayed effect), eventually affecting acceptance or
rejection of a sample.
}{%
}%

\iftoggle{full}{%
Clojure is a functional language, and continuation-passing style
(CPS) transformation is a well-developed technique in the area
of functional languages. Implementing a variant of CPS
transformation seemed to be the most flexible and lightweight
option --- any other form of concurrency would put a higher
burden on the underlying runtime (JVM) and the operating system.
Consequently, Anglican has been implemented as a CPS-transformed
computation with access to continuations in probabilistic
checkpoints. 
}{%
}%

@@

The operational semantics of an Anglican query are different from those
of Clojure code, and are determined by an inference algorithm. Thus,
Anglican queries must be called through these inference algorithms, rather
rather than `directly'. For this purpose,
\iftoggle{full}{the
\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
function \texttt{infer} using Clojure's multimethod mechanism.
This function accepts and runs an Anglican query, 
and returns a lazy sequence of weighted samples from the
distribution defined by the query. Providing an implementation of this function
is a responsibility of the inference algorithm, which should also
override the polymorphic function \texttt{checkpoint} (defined
as a multimethod) so as to handle \texttt{sample} and
\texttt{observe} in an algorithm-specific manner and
to construct an appropriate result on the
termination of a probabilistic program.

@@

@@

@@

;; invoked when algo parameter is :importance
(defmethod infer :importance 
  [algo prog value & {}]
  (letfn
    ;; recursive function without any parameter
    [sample-seq 
      []
      ;; lazy infinite sequence 
      (lazy-seq
        (cons
          (:state (exec ::algorithm prog value
                        initial-state))
          (sample-seq))))]
    (sample-seq)))
\end{lstlisting}
This implementation is dispatched 
when \texttt{infer} receives {\tt :im\-por\-tan\-ce} as its \texttt{algo} 
parameter. (Anglican includes multiple implementations of \texttt{infer}
for different inference algorithms.) Once dispatched, it lazily 
constructs an infinite sequence of inference states by repeatedly 
executing the program \texttt{prog} using \texttt{exec}
and retrieving the final inference state of the execution 
using \texttt{:state}. For checkpoint handlers, importance sampling simply
relies on their default implementations.

@@

\section*{Acknowledgements}

@@

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}
